Changes in trip-making frequency by mode during the COVID-19 emergency in Bangladesh

### 2020-10-23 Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
author: "Shaila Jamal and Antonio Paez"
date: "5/11/2020"
output: pdf_document
---

We wish to thank two anonymous reviewers for their comments and suggestions for revisions. In this letter we respond to those comments and explain how the paper was revised in response.

\newpage

# Reviewer 1

The paper presents an exploratory analysis of online survey results by 800 respondents on changes to their trip making behaviours by mode during the COVD pandemic. It is a well written paper, and I recommend publication. Some minor edits and adds are suggested below.

\textcolor{blue}{Thank you for your positive assessment and comments for improving the paper.}

- Figure 1: Please lengthen the y-axes for all figures, at the moment, it is a bit hard to read results off, because the axes are squashed.

\textcolor{blue}{Done. Hopefully the aspect of the figure will make it easier to read the results.}

- Table 1: Please mark whether the rows are before and columns are during in the cross tabulation. It would be a good idea to use something like a conditional colouring so that each cell value is coloured by the scale of the value, so that the patterns discussed come out visually clear. At the moment, one has to read the numbers carefully.

\textcolor{blue}{Words go here.}

- Otherwise, overall the paper is well written and the results are clearly discussed.

- Bangladesh has very high density urban areas, it would be good to include a discussion on whether the online survey responses indicated which cities the respondents are from - what was the geography of the answers? While 800 is a decent number, the volume is still a small number when compared as a sample against full populations, further, different cities have different transit arrangements, and since the shared and transit modes seem most affected - a discussion of geography somewhere (which principal cities and their attributes) will enrich the discussion, and also as mentioned a discussion of the limitations of the sample size being still relatively small.

\textcolor{blue}{Words go here.}

\newpage

# Reviewer 2

The level of reliability of the conclusions critically depends on the reliability of the sample, which is not analysed at all, i.e., there is no analysis of the level of representativeness of the sample relative to the Bangladesh population. A social media sample likely overrepresents high incomes, young age, etc. This must be scrutinized.

This is relevant because in the paper, from the title to the summary in the last paragraph, the authors claim they represent mobility changes in a developing country: Bangladesh, which has 160 million people (very small sample=800)

Other comments:

- No information of what contagion containment measures, if any, were in place during the time of the survey (e.g., lockdown, school closures)

- Table 1: columns and rows need to indicate which one is pre and during COVID, it has to be added to the table.

- Uber is not ridesharing, it is ridesourcing or ride-hailing.

\textcolor{blue}{This has been changed throughout the paper.}

### 2020-10-23 Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/Frequency-of-Travel-by-Mode-COVID-19-Bangladesh.Rmd
v2 Started working on revisions after reviews

---
title: An exploratory analysis of changes in trip-making frequency by mode during the COVID-19 emergency in Bangladesh
author:
  - name: Shaila Jamal
    email: jamals16@mcmaster.ca
    affiliation: McMaster University
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
 The COVID-19 pandemic has had a profound impact on mobility in every country and region around the world. Recent helps to illuminate some of the dimensions of change - however, the evidence is still scant in developing countries. The objective of this paper is to present an exploratory analysis of the changes in the frequency of trip-making by mode during the COVID-19 emergency in Bangladesh. The results confirm an overall loss of mobility in the form of reduced trip-making frequency by all modes, but the changes are not uniform across modes. In particular, there was a greater loss of mobility for bus, rickshaw, and CNG auto-rickshaw.

@@

header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

@@

# Pivot table to long: each row is a respondent-mode-period response
freq_by_mode_long <- freq_by_mode %>%
  pivot_longer(cols = starts_with("freq_"),
               names_to = "mode_period",
               names_prefix = "freq_",
               values_to = "frequency") %>%
  separate(mode_period, c("Mode", "Period")) %>%
  transmute(id,
            Period = factor(Period,
                            levels = c("before", "during")), 
            Mode = factor(Mode,
                          levels = c("car",
                                     "motorcycle",
                                     "rideshare",
                                     "rickshaw",
                                     "cng",
                                     "bus",
                                     "walk",
                                     "bicycle",
                                     "other")), 
            frequency = factor(frequency, 
                               levels = c("Never",
                                          "Rarely",
                                          "At least once a week",
                                          "2-3 times a week",
                                          "4-5 times a week",
                                          "Almost everyday"),
                               labels = c("Never",
                                          "Rarely",
                                          "Once a week",
                                          "2-3 per week",
                                          "4-5 per week",
                                          "Almost daily"),
                               ordered = TRUE))
```

@@

# RIDESHARE
rideshare_xtab <- xtabs(~ freq_rideshare_before + freq_rideshare_during, 
                        data = freq_by_mode)

@@

# RIDESHARE
rownames(rideshare_ptab2) = paste0("B_", all_states)
colnames(rideshare_ptab2) = paste0("D_", rev(all_states))

@@

For the purpose of this paper, we use two questions that provide information about trip-making frequency by eight modes of transportation. The modes are **car**, **ridesharing** (e.g., Uber, Pathao), **rickshaw**, **CNG auto-rickshaw** (a rickshaw-like vehicle powered by compressed natural gas), **bus**, motorcycle/scooter (hereafter just **motorcycle**), **walking**, and **bicycle** (there was an additional catch-all category **other** which we ignore here). Participants in the survey used the following levels to report their frequency of traveling by each mode both before and during COVID-19: _Never_, _Rarely_, _Once a week_, _2-3 trips per week_, _4-5 trips per week_, _Almost daily_. There are $n=800$ responses in the data set.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of 800) in each trip-making frequency class by mode of transportation. The white bars and gray bars are for travel before and during the pandemic, respectively. Considering travel before the pandemic, travel by rickshaw, and bus were relatively common for many respondents (few respondents reported _never_ using these modes). The mode most commonly used on a quotidian basis was walk. In contrast, respondents reported less frequent travel by car, ridesharing services, CNG auto-rickshaw, motorcycle, and bicycle. During the pandemic, we see that while there were reductions in mobility by car, motorcycle, and bicycle (with more respondents reporting never traveling by these modes), the changes were relatively minor.

```{r column-plot-cases-before-after, echo=FALSE, fig.width= 5, fig.cap="\\label{fig:column-plot-cases}Number of responses by trip-making frequency class and mode, before and during COVID-19"}
freq_by_mode_long %>%
  filter(Mode != "other") %>%
  ggplot(aes(x = frequency, y = Cases, fill = Period)) +
  geom_col(color = "black", position = "dodge") + 
  scale_fill_manual(values = c("before" = "white", "during" = "gray")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  facet_wrap(. ~ Mode, nrow = 4)
```

```{r table-transitions, echo=FALSE}
rbind(car_xtab, 
      rideshare_xtab,
      rickshaw_xtab, 
      cng_xtab,
      bus_xtab, 
      motorcycle_xtab,
      walk_xtab,
      bicycle_xtab) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>%
  pack_rows("car", 1, 6) %>%
  pack_rows("rideshare", 7, 12) %>%
  pack_rows("rickshaw", 13, 18) %>%
  pack_rows("CNG auto-rickshaw", 19, 24) %>%
  pack_rows("bus", 25, 30) %>%
  pack_rows("motorcycle", 31, 36) %>%
  pack_rows("walk", 37, 42) %>%
  pack_rows("bicycle", 43, 48)
```

The frequency of trip-making by other modes changed more noticeably: the frequency of travel by ridesharing services, rickshaw, CNG auto-rickshaw, and bus collapsed, with vastly more respondents reporting never using these modes during the pandemic than before. The frequency of walking trips also decreased (fewer respondents report walking almost daily), but the reductions in mobility were not so heavily concentrated at the bottom of the scale.

@@

From Figure \ref{fig:circular-plot-1}, we see that the probability of traveling less by car for those who initially used this mode frequently is high, but their probability of not using this mode at all during the pandemic is quite small. In other words, there was a decline in use, but not complete discontinuation of this mode. The probability of traveling more frequently by car for those who originally never or rarely used this mode remained is low. In contrast, we see that the probabilities of never ridesharing during the pandemic are high irrespective of the initial level of use of this mode of transportation, and the probability of using this mode more are fairly small.

```{r circular-plots-transition-probabilities-1, echo = FALSE, fig.cap="\\label{fig:circular-plot-1}Transition probabilities in trip frequency from before to during COVID-19: car and rideshare"}

@@

# RIDESHARE
circle.plot(rideshare_ptab2, "rideshare", nlines = -3)
```

@@

```{r circular-plots-transition-probabilities-2, echo = FALSE, fig.cap="\\label{fig:circular-plot-2}Transition probabilities in trip frequency from before to during COVID-19: rickshaw and cng auto-rickshaw"}
# Rows and columns to arrange several circular plots
par(mfrow = c(1, 2),
    mar = c(0, 0, 0, 0),
    oma = c(0, 0, 0, 0))

@@

# CNG
circle.plot(cng_ptab2, "CNG auto-rickshaw", nlines = -3)
```

After rideshare, the use of bus had the largest probabilities of discontinuation (Figure \ref{fig:circular-plot-3}). We see that even respondents who used this mode almost daily before the pandemic had close to 66% chances of never using it during the contingency. Similarly large losses of mobility by bus were observed for travelers who used this mode less frequently before the pandemic. These losses were not offset to any appreciable degree by the probability of some users turning to this mode during the pandemic.

```{r circular-plots-transition-probabilities-3, echo = FALSE, fig.cap="\\label{fig:circular-plot-3}Transition probabilities in trip frequency from before to during COVID-19: motorcycle and bus"}
# Rows and columns to arrange several circular plots
par(mfrow = c(1, 2),
    mar = c(0, 0, 0, 0),
    oma = c(0, 0, 0, 0))

@@

# MOTORCYCLE
circle.plot(motorcycle_ptab2, "motorcycle", nlines = -3)
```

@@

```{r circular-plots-transition-probabilities-4, echo = FALSE, fig.cap="\\label{fig:circular-plot-4}Transition probabilities in trip frequency from before to during COVID-19: bicycle and walk"}
# Rows and columns to arrange several circular plots
par(mfrow = c(1, 2),
    mar = c(0, 0, 0, 0),
    oma = c(0, 0, 0, 0))

@@

# BICYCLE
circle.plot(bicycle_ptab2, "bicycle", nlines = -3)
```

The findings reported here provide important changes in mobility in a developing country during COVID-19. In general there was a loss of mobility across-the-board, but modes that require interaction with strangers (rideshare, bus, rickshaw, and CNG auto-rickshaw) were more affected. Travel by car, motorcycle, and bicycle were somewhat less affected, and there is evidence of some adoption of walking during the pandemic.

>>>>

---
title: Changes in trip-making frequency by mode during the COVID-19 emergency in Bangladesh
author:
  - name: Shaila Jamal
    email: jamals16@mcmaster.ca
    affiliation: McMaster University
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
 The COVID-19 pandemic has had a profound impact on mobility in every country and region around the world. Recent helps to illuminate some of the dimensions of change - however, the evidence is still scant in developing countries. The objective of this paper is to present an exploratory analysis of the changes in the frequency of trip-making by mode during the COVID-19 emergency in Bangladesh. The results confirm an overall loss of mobility in the form of reduced trip-making frequency by all modes, but the changes are not uniform across modes. In particular, there was a greater loss of mobility for bus, rickshaw, and CNG auto-rickshaw, and some adoption of walking during the pandemic.

@@

header-includes:
 \usepackage{float}
 \usepackage{xcolor}
 \floatplacement{figure}{H}

---

@@

# Pivot table to long: each row is a respondent-mode-period response
freq_by_mode_long <- freq_by_mode %>%
  pivot_longer(cols = starts_with("freq_"),
               names_to = "mode_period",
               names_prefix = "freq_",
               values_to = "frequency") %>%
  separate(mode_period, c("Mode", "Period")) %>%
  transmute(id,
            Period = factor(Period,
                            levels = c("before", "during")), 
            Mode = factor(Mode,
                          levels = c("car",
                                     "motorcycle",
                                     "rideshare",
                                     "rickshaw",
                                     "cng",
                                     "bus",
                                     "walk",
                                     "bicycle",
                                     "other"),
                          labels = c("car",
                                     "motorcycle",
                                     "ride-hailing",
                                     "rickshaw",
                                     "cng",
                                     "bus",
                                     "walk",
                                     "bicycle",
                                     "other")), 
            frequency = factor(frequency, 
                               levels = c("Never",
                                          "Rarely",
                                          "At least once a week",
                                          "2-3 times a week",
                                          "4-5 times a week",
                                          "Almost everyday"),
                               labels = c("Never",
                                          "Rarely",
                                          "Once a week",
                                          "2-3 per week",
                                          "4-5 per week",
                                          "Almost daily"),
                               ordered = TRUE))
```

@@

# RIDE-HAILING
rideshare_xtab <- xtabs(~ freq_rideshare_before + freq_rideshare_during, 
                        data = freq_by_mode)

@@

# RIDE-HAILING
rownames(rideshare_ptab2) = paste0("B_", all_states)
colnames(rideshare_ptab2) = paste0("D_", rev(all_states))

@@

For the purpose of this paper, we use two questions that provide information about trip-making frequency by eight modes of transportation. The modes are **car**, **ride-hailing** (e.g., Uber, Pathao), **rickshaw**, **CNG auto-rickshaw** (a rickshaw-like vehicle powered by compressed natural gas), **bus**, motorcycle/scooter (hereafter just **motorcycle**), **walking**, and **bicycle** (there was an additional catch-all category **other** which we ignore here). Participants in the survey used the following levels to report their frequency of traveling by each mode both before and during COVID-19: _Never_, _Rarely_, _Once a week_, _2-3 trips per week_, _4-5 trips per week_, _Almost daily_. There are $n=800$ responses in the data set.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of 800) in each trip-making frequency class by mode of transportation. The white bars and gray bars are for travel before and during the pandemic, respectively. Considering travel before the pandemic, travel by rickshaw, and bus were relatively common for many respondents (few respondents reported _never_ using these modes). The mode most commonly used on a quotidian basis was walk. In contrast, respondents reported less frequent travel by car, ride-hailing services, CNG auto-rickshaw, motorcycle, and bicycle. During the pandemic, we see that while there were reductions in mobility by car, motorcycle, and bicycle (with more respondents reporting never traveling by these modes), the changes were relatively minor.

```{r column-plot-cases-before-after, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:column-plot-cases}Number of responses by trip-making frequency class and mode, before and during COVID-19"}
freq_by_mode_long %>%
  filter(Mode != "other") %>%
  ggplot(aes(x = frequency, y = Cases, fill = Period)) +
  geom_col(color = "black", position = "dodge") + 
  scale_fill_manual(values = c("before" = "white", "during" = "gray")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  facet_wrap(. ~ Mode, nrow = 4)

# Uncomment to save to jpg files for the publisher
#ggsave(filename = "Figure-1.jpeg", device = "jpeg", height = 9, width = 6.5, units = "in")
```

```{r table-transitions, echo=FALSE}
rbind(car_xtab, 
      rideshare_xtab,
      rickshaw_xtab, 
      cng_xtab,
      bus_xtab, 
      motorcycle_xtab,
      walk_xtab,
      bicycle_xtab) %>% 
  t() %>%
  data.frame() %>% 
  mutate(across(.cols = everything(), function(x) {
    cell_spec(x, "latex", bold = TRUE, color = "white", background = spec_color(x, end = 0.8, alpha = 0.3, option = "plasma", direction = -1))
    })) %>%
  t() %>%
  data.frame() %>%
  rename(Never = X1, Rarely = X2, `Once a week` = X3, `2-3 times per week` = X4, `4-5 times per week` = X5, `Almost daily` = X6) %>%
  rownames_to_column(var = "Before") %>%
  mutate(Before = rep(c("Never", "Rarely", "Once a week", "2-3 times per week", "4-5 times per week", "Almost daily"), 8)) %>%
  kable("latex",
        booktabs = TRUE,
        escape = FALSE,
        caption = "\\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>%
  add_header_above(c(" " = 1, "During" = 6)) %>%
  pack_rows("car", 1, 6) %>%
  pack_rows("ride-hailing", 7, 12) %>%
  pack_rows("rickshaw", 13, 18) %>%
  pack_rows("CNG auto-rickshaw", 19, 24) %>%
  pack_rows("bus", 25, 30) %>%
  pack_rows("motorcycle", 31, 36) %>%
  pack_rows("walk", 37, 42) %>%
  pack_rows("bicycle", 43, 48)  %>%
  footnote(general = "Darker cell colors indicate higher cell values in a row.")
```

The frequency of trip-making by other modes changed more noticeably: the frequency of travel by ride-hailing services, rickshaw, CNG auto-rickshaw, and bus collapsed, with vastly more respondents reporting never using these modes during the pandemic than before. The frequency of walking trips also decreased (fewer respondents report walking almost daily), but the reductions in mobility were not so heavily concentrated at the bottom of the scale.

@@

From Figure \ref{fig:circular-plot-1}, we see that the probability of traveling less by car for those who initially used this mode frequently is high, but their probability of not using this mode at all during the pandemic is quite small. In other words, there was a decline in use, but not complete discontinuation of this mode. The probability of traveling more frequently by car for those who originally never or rarely used this mode remained is low. In contrast, we see that the probabilities of never ride-hailing during the pandemic are high irrespective of the initial level of use of this mode of transportation, and the probability of using this mode more are fairly small.

```{r circular-plots-transition-probabilities-1, echo = FALSE, fig.cap="\\label{fig:circular-plot-1}Transition probabilities in trip frequency from before to during COVID-19: car and ride-hailing"}
# Uncomment to save to jpg files for the publisher
#jpeg("Figure-2.jpg", width = 6.5, height = 4.5, units = "in", res = 600)

@@

# RIDE-HAIL
circle.plot(rideshare_ptab2, "ride-hailing", nlines = -3)

# Uncomment to save to jpg files for the publisher
#dev.off()
```

@@

```{r circular-plots-transition-probabilities-2, echo = FALSE, fig.cap="\\label{fig:circular-plot-2}Transition probabilities in trip frequency from before to during COVID-19: rickshaw and cng auto-rickshaw"}
# Uncomment to save to jpg files for the publisher
#jpeg("Figure-3.jpg", width = 6.5, height = 4.5, units = "in", res = 600)

@@

# CNG
circle.plot(cng_ptab2, "CNG auto-rickshaw", nlines = -3)

# Uncomment to save to jpg files for the publisher
#dev.off()
```

After ride-hailing, the use of bus had the largest probabilities of discontinuation (Figure \ref{fig:circular-plot-3}). We see that even respondents who used this mode almost daily before the pandemic had close to 66% chances of never using it during the contingency. Similarly large losses of mobility by bus were observed for travelers who used this mode less frequently before the pandemic. These losses were not offset to any appreciable degree by the probability of some users turning to this mode during the pandemic.

```{r circular-plots-transition-probabilities-3, echo = FALSE, fig.cap="\\label{fig:circular-plot-3}Transition probabilities in trip frequency from before to during COVID-19: motorcycle and bus"}
# Uncomment to save to jpg files for the publisher
#jpeg("Figure-4.jpg", width = 6.5, height = 4.5, units = "in", res = 600)

@@

# MOTORCYCLE
circle.plot(motorcycle_ptab2, "motorcycle", nlines = -3)

# Uncomment to save to jpg files for the publisher
#dev.off()
```

@@

```{r circular-plots-transition-probabilities-4, echo = FALSE, fig.cap="\\label{fig:circular-plot-4}Transition probabilities in trip frequency from before to during COVID-19: bicycle and walk"}
# Uncomment to save to jpg files for the publisher
#jpeg("Figure-5.jpg", width = 6.5, height = 4.5, units = "in", res = 600)

@@

# BICYCLE
circle.plot(bicycle_ptab2, "bicycle", nlines = -3)

# Uncomment to save to jpg files for the publisher
#dev.off()
```

The findings reported here provide important changes in mobility in a developing country during COVID-19. In general there was a loss of mobility across-the-board, but modes that require interaction with strangers (ride-hailing, bus, rickshaw, and CNG auto-rickshaw) were more affected. Travel by car, motorcycle, and bicycle were somewhat less affected, and there is evidence of some adoption of walking during the pandemic.
